## Reproducible Research: Peer Assessment 2

# Analysis of Health and Economic Impacts of Severe Weather Events Across the United States

Sagi Greenstine

## Assignment

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

## Synopsis

This analysis based on the U.S. National Oceanic and Atmospheric Administration's (NOAA) Storm Database, that contains the events starting from 1950 until November 2011. 
The NOAA Storm Database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
This data analysis project tries to answer the following questions:
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The conclusions of this analysis are:
1. The most harmful events to human health in terms of fatality and injuries are: tornado, excessive heat, flash flood, thunderstorms, cold etc.
2. The types of events which have the greatest economic consequences in terms of property and crop damages are: various kinds of strong winds, floods, fires, relatively crop damages is also cold. etc.

## Data Processing

### Loading and preprocessing the data

```{r loaddata}
if(!exists("stormdata")){
  stormdata <- read.csv("./repdata-data-StormData.csv.bz2", row.names = NULL, nrows=902297, check.names=F, stringsAsFactors=F, header=T, sep=',', na.strings="NA", comment.char="", quote='\"')
}
```

```{r strdataset, results="hide""}
str(stormdata)
```

This dataset has 902,297 observations of  37 variables.

Also, one of the important parameteres is event type (EVTYPE variable). The values of event types don't have a specific format.

I have to clean and tidy the data before I answer the questions.

Let see a number of unique values of EVTYPE in raw dataset:

```{r uniquerawdata}
length(unique(stormdata$EVTYPE))
```

There are 985 unique values of event types. Let's try to reduce this amount by means of the following operations:

1. Bringing all the values of EVTYPE to the lowercase format:

```{r lowercase}
tidyevtype <- tolower(stormdata$EVTYPE)
```

2. Replacement of punctuation, control and blank characters with the space

```{r repcharacters}
tidyevtype <- gsub("[[:blank:][:punct:][:cntrl:]+]", " ", tidyevtype)
length(unique(tidyevtype))
```

Now, there are 874 unique values of event types.

Let's update the dataset with processed values of event types:

```{r evtypeupdate}
stormdata$EVTYPE <- tidyevtype
```

3. Reducing of unique similar values of event types to similar instances:

```{r similarevtype}
damagetype <- stormdata$EVTYPE
damagetype <- gsub("burst|depression|drizzle|fog|hail|percip|precipitation|rain|wall cloud|wet", "Precipitations", damagetype)
damagetype <- gsub("hurricane|storm|typhoon|wind|wnd", "Winds", damagetype)
damagetype <- gsub("erosion|slide|slump", "Landslides", damagetype)
damagetype <- gsub("dry|drought|heat|hot|record high|record temperature|temperature record|thermia|warmth|warm", "Heat", damagetype)
damagetype <- gsub("avalanche|blizzard|chill|cold|cool|freezing|freeze|frost|glaze|ice|icy|sleet|snow|winter|wintery|wintry", "Cold", damagetype)
damagetype <- gsub("blow-out|dam break|fld|flood|surf|swells", "Flood", damagetype)
damagetype <- gsub("current|drowing|high water|marine|seas|tide|tsunami|wave", "Waves", damagetype)
damagetype <- gsub("dust|saharan", "Dust", damagetype)
damagetype <- gsub("lightning|thunderstorm|tstm", "Thunderstorms", damagetype)
damagetype <- gsub("funnel|tornado|spout|whirlwind", "Tornado", damagetype)
damagetype <- gsub("fire|smoke|volcanic", "Fire", damagetype)
stormdata <- cbind(stormdata, damagetype)
length(unique(damagetype))
```

Now, there are 774 unique values of event types.

4. Now I need to convert char. variables PROPDMGEXP (property damage extension) and CROPDMGEXP (crop damage extension) into the corresponding numeric values, for example, "h" = 100, "k" = 1,000, "M" = 1,000,000. For these needs, I'll use the following function:

```{r extconvert}
extconvert <- function(x) {
                  if (x %in% c('h', 'H')) 
                      x <- 2 
                   else  if (x %in% c('k', 'K')) 
                      x <- 3 
                   else  if (x %in% c('M', 'm')) 
                      x <- 6 
                   else  if (x %in% c('B', 'b')) 
                      x <- 9
                   else  if (x == "" || x == " ")
                      x <- 0
                   else  
                      x <- NA
                      x
}                 
```

Converting the variables and computing the absolute values of damages:

```{r absdamages}
abspropdmg <- sapply(stormdata$PROPDMGEXP, FUN=extconvert)
stormdata$abspropdamage <- stormdata$PROPDMG * (10 ** abspropdmg)
abscropdmg <- sapply(stormdata$CROPDMGEXP, FUN=extconvert)
stormdata$abscropdamage <- stormdata$CROPDMG * (10 ** abscropdmg)
```

## Results

Now, after the loading and preprocessing the data, let's answer on following questions.

#### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Let's find the most harmful event types for population health, across the US.

```{r healthdamage}
library(plyr)
healthdmg <- ddply(stormdata, .(damagetype), summarize, sumfatality = sum(FATALITIES), suminjury = sum(INJURIES))
fatalityev <- head(healthdmg[order(healthdmg$sumfatality, decreasing = T), ], 10)
injuryev <- head(healthdmg[order(healthdmg$suminjury, decreasing = T), ], 10)
```

Top 10 event types that caused largest number of fatalities:

```{r topfatalities}
fatalityev[, c("damagetype", "sumfatality")]
```

Top 10 event types that caused largest number of injuries:

```{r topinjuries}
injuryev[, c("damagetype", "suminjury")]
```

Now I want to show the most harmful types of events to public health through the following chart:

```{r healthplot}
library(ggplot2)
library(gridExtra)
fatalplot <- ggplot(data=fatalityev, aes(x=reorder(damagetype, sumfatality), y=sumfatality)) + geom_bar(stat="identity") +
coord_flip() + ylab("Total deaths") + xlab("Event Type") + theme(legend.position="none")

injuryplot <- ggplot(data=injuryev, aes(x=reorder(damagetype, suminjury), y=suminjury)) + geom_bar(stat="identity") +
coord_flip() + ylab("Total injuries") + xlab("Event Type") + theme(legend.position="none")

grid.arrange(fatalplot, injuryplot)
```

Based on the above data samples and graphs, you can see that the most harmful events to human health in terms of fatality and injuries are: tornado, excessive heat, flash flood, thunderstorms, cold etc.

#### 2. Across the United States, which types of events have the greatest economic consequences?

Let's see,  which types of events have the greatest economic consequences, across the US, in terms of property and corp damages.

```{r economicdamage}
economicdmg <- ddply(stormdata, .(damagetype), summarize, sumpropdamage = sum(abspropdamage), sumcropdamage = sum(abscropdamage))
propertydmg <- head(economicdmg[order(economicdmg$sumpropdamage, decreasing = T), ], 10)
cropdmg <- head(economicdmg[order(economicdmg$sumcropdamage, decreasing = T), ], 10)
```

Top 10 event types that caused largest property damage:

```{r toppropertydmg}
propertydmg[, c("damagetype", "sumpropdamage")]
```

Top 10 event types that caused largest crop damage:

```{r topcropdmg}
cropdmg[, c("damagetype", "sumcropdamage")]
```

Now, let's see the event types that have the greatest economic consequences through the following chart:

```{r economicplot}
propdmgplot <- ggplot(data=propertydmg, aes(x=reorder(damagetype, sumpropdamage), y=sumpropdamage)) + geom_bar(stat="identity") + coord_flip() +     ylab("Total property damage (in USD)") + xlab("Event Type") + theme(legend.position="none")

cropdmgplot <- ggplot(data=cropdmg, aes(x=reorder(damagetype, sumcropdamage), y=sumcropdamage)) + geom_bar(stat="identity") + coord_flip() + 
xlab("Event type") + ylab("Total crop damage (in USD)") + theme(legend.position="none")

grid.arrange(propdmgplot, cropdmgplot)
```

Based on the above data samples and graphs, you can see that the types of events which have the greatest economic consequences in terms of property and crop damages are: various kinds of strong winds, floods, fires, relatively crop damages is also cold. etc.
